Exploring the composition of lithic assemblages in Mesolithic south-eastern Norway

### 2021-05-17 analysis/paper/paper.Rmd
Writing

This study employs multivariate exploratory statistics to analyse the entirety of the lithic assemblages associated with a larger number of Mesolithic sites located in south-eastern Norway. This is done to identify latent patterns and structure in the relationship between the assemblages, with the ultimate aim of identifying behaviourally induced variation in their composition across time. However, the composition of the assemblages can be expected to be determined by a multitude of factors [e.g. @dibble2017; @rezek2020], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency and group sizes at visits to the sites. The assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials. Finally, analytical and methodological dimensions relating to survey, excavation and classification practices are also fundamental to how the assemblages are defined. Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern. In an attempt to limit the influence of confounding effects, the material chosen for analysis has a relatively constrained geographical distribution, and stems from relatively recent investigations that have employed comparable methods for excavation and classification.

@@

The data analysed here is based on the classification of the `r nrow(sites)` site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types. While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions. Some categories in the original reports have therefore been combined in the dataset. This for example pertains to the category narrow-blades, which is defined as blades of width between 8 and 12 mm. In the reports, this was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites. Narrow-blades were combined with the blade category here. Furthermore, the artefact data have here been divided into flint and non-flint materials. Flint does not occur indigenously in Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of the flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [@eigeland2015, 46]. An in depth-analysis of the flint quality from sevens sites did result in the finding that the flint quality was relatively high, and consistently so across the sites. However, despite flint being treated as a unified category here, the cited study by was based on only seven sites located some km distance from the present study area, meaning that the results might not necessarily be transferable to the sites treated here [@eigeland2015, 45-53]. Furthermore, the various non-flint raw materials that have been lumped together here can have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories. An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf. @frivoll2017]. While certainly a topic deserving of more attention, the general sentiment in the literature is that there would have been stable access to locally available non-flint raw-materials of good quality in south-eastern Norway [@eigeland2015, 370].

@@

As the title of the paper suggest, the data has been analysed from an exploratory perspective, meaning that a wide range of methods has been applied, and various combinations and transformations of variables has been explored. Only parts of this process can sensibly be reported. However, this kind of data dredging does arguably not constitute a good foundation for the pursuit of causal relationships, as this process is most certainly selectively reported upon and heavily impacted by my own biases. It can, however, result in the identification of interesting relationships that can be subjected to more rigorous testing using different data and approaches in subsequent studies. Furthermore, the employed data and R script is freely available as a research compendium following @marwick, allowing readers to explore and scrutinise the data and the final analytical choices made [cf. @Marwick2017].

@@

Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for mobility patterns as operationalised by @clark2017 and @Bicho2020. There is a general temporal transition from older to younger sites along the second dimension of the plot, which is mainly defined by a negative correlation between the variables volumetric density of lithics and relative frequency of secondarily worked lithics, as well as relative frequency of cores. Almost orthogonal to this is the negative correlation between relative frequency of chips and blanks. While there is perhaps a slight tendency for blanks to be more associated with younger sites, and chips with older sites, this relationship does in large part appear to be temporally determined. The investigation performed by @Bicho2020 indicated that in their data the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns. However, they also found that the relationship between volumetric density of lithics and the frequency of retouched artefacts gave a reasonable estimation of mobility, which, combined with the fact that the relative frequency of cores align with the expectation from the @clark2017 model, would indicate that this relationship is capturing the dimensions of interest here. If the relationship between frequency of secondarily worked artefacts and volumetric density of lithics is accepted as a proxy for curation, and in turn related to land-use and mobility patterns, these findings would be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases. To explore this proposition further, these two variables are subjected to more detailed scrutiny below.

@@

Figure \@ref(fig:wabi) displays the WABI as found for the site data. This indicates a negative correlation between the two variables (r = `r round(wabicor, 2)`) and a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites. Following @clark2017, this would be in line with a general transition from a expedient to curated lithic technology. Variable raw-material availability can also impact these measures as [@manninen2014]. Variation in raw-material use over time could consequently mean that these patterns could be an effect of the available raw material rather than mobility strategies. However, as is evident in when considering both plots in \@ref(fig:wabi), the pattern is evident for both the lithics considered in aggregate and for the flint data specifically. Although the distribution is slightly more spread out along the y-axis than the x-axis in the flint data, indicating that differences in volumtric density of flint is not as clearly contingent on chronology, the general pattern is the same.

@@

To visualise this temporal trend more clearly, a curation index based on the two variables was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1. The values for artefact density was then made negative to reflect its relationship with degree of curation. The mean was then found for each site on these two normalised values. To account for the temporal uncertainty associated with the dating of the sites, a simulation-based approach was also adopted [e.g. @crema2012; @baxter2016; @orton2017]. A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from the date ranges associated with the sites. For sites with radiocarbon age determinations the dates were drawn from the associated posterior probability density functions, while age for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range. This simulation was repeated 1000 times, the results of which is visualised in \@ref(fig:simage). Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE. This stabilises around 7000 BCE without any further major fluctuations for the rest of the Mesolithic. It is also interesting to note that the variation in degree of curation is higher after 8000 BCE, potentially reflecting variation in site activities and associated mobility patterns.

@@

Assuming for now that the curation index captures degree of mobility, the present data indicates a generally higher degree of mobility until some time before 8000 BCE. This then drops and stabilises from around 7000 BCE for the rest of the Mesolithic. This pattern is evident in both the flint data and when all lithics are treated in aggregate. The variation in degree of curation when all lithics are considered, could indicate that the sites were associated with a more varied mobility pattern after around 8000 BCE. The five sites that have values on the curation index below c. -0.25 could in this perspective have predominantly functioned as base camps within a logistic settlement pattern. That these assemblages reflect stays of a longer duration was suggested for all five sites in the original reports [@solheim2013; @carrasco2014; @eigeland2014b; @persson2014; @eigeland2017], with the exception of for Vallermyrene 4, which was argued to be a specialised axe production site, not necessarily associated with lower degrees of mobility. This highlights a possible issue pertaining to raw-material variability, as the coarse non-flint material used for the production of axes generally results in a relatively large amount of waste per produced tool, possibly skewing the curation index when compared to assemblages dominated by flint. Also referring back to the CA, the difference is most marked for the sites in the later part of the Mesolithic where non-flint material become more dominating parts of the assemblages. As can be seen on the plot only using the flint-data, the degree of curation is markedly higher for both Gunnarsrød 6b and Vallermyrene 4 when the non-flint material is excluded, although they remain more expedient than that of contemporary assemblages. The patterns evident in the flint data could thus indicate that only the period from around 8000 to 7500 BCE has individual sites that have had more stable roles within a settlement system, where some predominantly have functioned as base-camps. The lower degree of variation in the proceeding and preceeding phases could reflect either a more uniform purpose of visits to the sites, or that each assemblage consitutes the aggregation of visits of different purposes that in effect cancel each other out. Given the overall higher expediency in the earliest part of the Mesolithic this could be in line with a residental mobility pattern. For the later Mesolithic, the high expediency on some sites when considering all lithics in aggregate does speak to variability in site use, but the nature of the non-flint material does leave the signficance of this pattern more difficult to directly align with the residential logistic continuum. A case also worth commenting on is Langangen Vestgård 1, which, on the grounds of a overall large number of artefacts and the possible presence of a dwelling structure was argued to reflect a more permanent site location in the original report [@melvold2014]. However, the relatively high value on the curation index could mean that Langangen Vestgård 1 reflects the aggregation of stays which predominantly have been of a relatively shorter duration, while the possible dwelling structure, if taken as an indication of longer stays, could represent a remnant from one or a few visits of longer duration that constitute a smaller fraction of the use-life of the site as a whole.

>>>>

This study employs multivariate exploratory statistics to analyse the entirety of the lithic assemblages associated with a larger number of Mesolithic sites located in south-eastern Norway. This is done to identify latent patterns and structure in the relationship between the assemblages, with the ultimate aim of identifying behaviourally induced variation in their composition across time. However, the composition of the assemblages can be expected to be determined by a multitude of factors [e.g. @dibble2017; @rezek2020], ranging from the impact of natural formation processes, to various and intermixed behavioural aspects such as purpose, duration, frequency and group sizes at visits to the sites. The assemblages are also likely to be impacted by variation in lithic technology, artefact function, use-life and discard patterns, as well as access to raw materials. Finally, analytic and methodological dimensions relating to survey, excavation and classification practices are also fundamental to how the assemblages are defined. Consequently, the analysis conducted here is done from an exploratory perspective, where all of these factors should be seen as potential contributors to any observed pattern. In an attempt to limit the influence of confounding effects, the material chosen for analysis has a relatively constrained geographical distribution, and stems from relatively recent investigations that have employed comparable methods for excavation and classification.

@@

The data analysed here is based on the classification of the `r nrow(sites)` site assemblages done for the original excavation reports, and consists of `r ncol(select(sites, core:flake_retouch_other))` variables representing different debitage and tool types. While the classification practices for the excavation projects were standardised to an extent, there are some instances where time was allocated to identify additional artefact sub-categories aimed at answering specific research questions. Some categories in the original reports have therefore been combined in the dataset. This for example pertains to the category narrow-blades, which is defined as blades of width between 8 and 12 mm. In the reports, this was only separated from (macro-)blades (width $\ge$ 12 mm) and micro-blades (width $\le$ 8 mm) for some of the sites. Narrow-blades were combined with the blade category here. Furthermore, the artefact data have here been divided into flint and non-flint materials. Flint does not occur indigenously in Norway, and is only available locally as nodules that have been transported and deposited by retreating and drifting ice [e.g. @berg-hansen1999]. This means that the distribution and quality of the flint has been impacted by a diverse set of factors relating to climatic and geographical factors such as, but not limited to, topographic variability, shoreline morphology and ocean currents [@eigeland2015, 46]. An in depth-analysis of the flint quality from sevens sites did result in the finding that the flint quality was relatively high, and consistently so across the sites. However, despite flint being treated as a unified category here, the cited study by was based on only seven sites located some km distance from the present study area, meaning that the results might not necessarily be transferable to the sites treated here [@eigeland2015, 45-53]. Furthermore, the various non-flint raw materials that have been lumped together here can have quite disparate properties, where fine-grained cryptocrystalline materials are often used as a substitute or supplement to flint, while other, coarser materials are usually associated with the production of axes and other macro tools. Given this differentiated use, these raw-material properties are expected to be reflected in the retained debitage and tool categories. An important benefit of combining all of the non-flint materials is that this reduces the dependency on whether or not these have been correctly and consistently categorised for the reports [cf. @frivoll2017]. While certainly a topic deserving of more attention, the general sentiment in the literature is that there would have been stable access to locally available non-flint raw-materials of good quality in south-eastern Norway [e.g. @glørstad2011; @eigeland2015, 370].

@@

As the title of the paper suggest, the data has here been analysed from an exploratory perspective, meaning that a wide range of methods has been applied, and various combinations and transformations of variables has been explored. Only parts of this process can sensibly be reported. However, this kind of data dredging does arguably not constitute a good foundation for the pursuit of causal relationships, as this process is most certainly selectively reported upon and heavily impacted by my own biases. It can, however, result in the identification of interesting relationships that can be subjected to more rigorous testing using different data and approaches in subsequent studies. Furthermore, the employed data and R script is freely available as a research compendium following @marwick, allowing readers to explore and scrutinise the data and the final analytical choices made [cf. @Marwick2017].

@@

Figure \@ref(fig:pca-analysis) displays a principle components analysis using variations of the continuous measures for mobility patterns as operationalised by @clark2017 and @Bicho2020. There is a general temporal transition from older to younger sites along the second dimension of the plot, which is mainly defined by a negative correlation between the variables volumetric density of lithics and relative frequency of secondarily worked lithics, as well as relative frequency of cores. Almost orthogonal to this is the negative correlation between relative frequency of chips (artefacts of a size \>) and blanks (flakes, blades, microblades and fragments). While there is perhaps a slight tendency for blanks to be more associated with younger sites, and chips with older sites, this relationship does in large part appear not to be temporally determined. The investigation performed by @Bicho2020 indicated that in their data the relative frequency of chips, cores, and blanks might be more sensitive to mobility patterns. However, they also found that the relationship between volumetric density of lithics and the frequency of retouched artefacts gave a reasonable estimation of mobility, which, when combined with the fact that the relative frequency of cores align with the expectation of the @clark2017 model, would indicate that this relationship is capturing the dimensions of interest here. If the relationship between frequency of secondarily worked artefacts and volumetric density of lithics is accepted as a proxy for curation, and is related to land-use and mobility patterns, these findings would be in line with previous research into the Mesolithic of Norway, indicating that earlier sites are associated with higher degree of mobility than sites from later phases. To explore this proposition further, these two variables are subjected to more detailed scrutiny below.

@@

Figure \@ref(fig:wabi) clearly illustrates the negative correlation between the two variables (r = `r round(wabicor, 2)`) and a general tendency for younger sites to be associated with a higher volumetric density of lithics and a lower relative frequency of secondarily worked lithics than older sites. Following @clark2017, this would be in line with a general transition from a expedient to curated lithic technology. As variable raw-material availability and workability can also impact these measures [@manninen2014]. Variation in raw-material use over time could consequently mean that these patterns could be an effect of the available raw material rather than mobility strategies. However, as is evident in when considering both plots in \@ref(fig:wabi), the pattern is evident for both the lithics considered in aggregate and for the flint data specifically. Although the distribution is slightly more spread out along the y-axis than the x-axis in the flint data, indicating that differences in volumtric density of flint is not as clearly temporally contingent, the general pattern is the same.

@@

To visualise this temporal trend more clearly, a curation index based on the two variables was devised by first performing a min-max normalisation of the two variables, scaling them to take on values between 0 and 1. The values for artefact density was then made negative to reflect its relationship with degree of curation. The mean was then found for each site on these two normalised values. To account for the temporal uncertainty associated with the dating of the sites, a simulation-based approach was also adopted [e.g. @crema2012; @baxter2016; @orton2017]. A LOESS curve was fit to the curation index and site age for each simulation run, where the age of each site was drawn as a single year from the date ranges associated with the sites \@ref(fig:dist). For sites with radiocarbon age determinations the dates were drawn from the associated posterior probability density functions, while ages for sites dated with reference to relative sea-level change and typology were drawn uniformly from the associated date range. This simulation was repeated 1000 times, the results of which is visualised in \@ref(fig:simage). Disregarding the edge-effects at either end of the plot, the general tendency is a relatively high degree of curation among the earlier sites, followed by a marked drop around 8000 BCE. This stabilises around 7000 BCE without any further major fluctuations for the rest of the Mesolithic. It is also interesting to note that the variation in degree of curation is higher after 8000 BCE, potentially reflecting variation in site activities and associated mobility patterns.

@@

The curation index has relatively high values until some time before 8000 BCE, before it drops and stabilises from around 7000 BCE for the rest of the Mesolithic. This pattern is evident in both the flint data and when all lithics are treated in aggregate. The variation in degree of curation when all lithics are considered, could indicate that the sites were associated with a more varied mobility pattern after around 8000 BCE. The five sites that have values on the curation index below c. -0.25 could in this perspective have predominantly functioned as base camps within a logistic settlement pattern. That these assemblages reflect stays of a longer duration was suggested for all five sites in the original reports [@solheim2013; @carrasco2014; @eigeland2014b; @persson2014; @eigeland2017], with the exception of for Vallermyrene 4, which was argued to be a specialised axe production site, not necessarily associated with lower degrees of mobility. This highlights a possible issue pertaining to raw-material variability, as the coarse non-flint material used for the production of axes generally results in a relatively large amount of waste per produced tool, possibly skewing the curation index when compared to assemblages dominated by flint. Also referring back to the CA, the difference is most marked for the sites in the later part of the Mesolithic where non-flint material become more dominating parts of the assemblages. As can be seen on the plot only using the flint-data, the degree of curation is markedly higher for both Gunnarsrød 6b and Vallermyrene 4 when the non-flint material is excluded, although they remain more expedient than that of contemporary assemblages. The patterns evident in the flint data could thus indicate that only the period from around 8000 to 7500 BCE has individual sites that have had more stable roles within a settlement system, where some predominantly have functioned as base-camps. The lower degree of variation in the proceeding and preceeding phases could reflect either a more uniform purpose of visits to the sites, or that each assemblage consitutes the aggregation of visits of different purposes that in effect cancel out. Given the overall higher expediency in the earliest part of the Mesolithic this could be in line with a residental mobility pattern, while the high expediency on some sites when considering all lithics in aggregate in the later parts of the Mesolithic does speak to variability in site use. The nature of the non-flint material does leave the signficance of this pattern more difficult to directly align with the residential-logistic continuum. A case also worth commenting on is Langangen Vestgård 1, which, on the grounds of a overall large number of artefacts and the possible presence of a dwelling structure was argued to reflect a more permanent site location in the original report [@melvold2014]. However, the relatively high value on the curation index could mean that Langangen Vestgård 1 reflects the aggregation of stays which predominantly have been of a relatively shorter duration, while the possible dwelling structure, if taken as an indication of longer stays, could represent a remnant from one or a few visits of longer duration that constitute a smaller fraction of the use-life of the site as a whole.
